---
title: "Baseline Characteristics Table (Table 1) — Full Pipeline"
author: "Saad Ishaq Khattak"
date: "Confirmed running on Barkla HPC, 12 July 2026 (prep job 9762235, baseline table job 9762366)"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(eval = FALSE, echo = TRUE)
```

# Purpose

This document is a combined record of the full pipeline behind Table 1:
cohort construction (Steps 1-7) followed by the table itself (Steps 8-9).

**On Barkla, these ran as two separate scripts, not one file:**
`prep_cohort.R` (Steps 1-7) produces `year1.rds` and `pwp.rds`, saved to
`/users/hlskhatt/outputs/`. `baseline_table.R` (Steps 8-9) loads those
two files and does not rebuild the cohort itself. This split is
deliberate — cohort-construction logic lives in exactly one place, so a
fix only ever needs making once, and the Noble validation script reuses
the same `year1.rds`/`pwp.rds` rather than duplicating Steps 1-7 again.

This combined file is a documentation convenience, not a literal
single-script transcript. `eval=FALSE` throughout — nothing in this
document executes when rendered. No patient data or output files are
included in this repository — code only.

**Dependency notes:**
- Steps 1-7 need `dplyr`, `data.table`, `moments`, `lubridate`, `ggplot2`.
- Steps 8-9 need `dplyr`, `gtsummary`, and a working `pandoc`. Barkla has
  no system-wide pandoc; a standalone binary was downloaded to
  `~/pandoc_local/pandoc-3.1.11/bin/pandoc` and added to `PATH` rather
  than installing `flextable`/`officer`/`ragg`, which failed to compile
  due to missing `freetype`/`libtiff` system libraries on this cluster.

**Status:** confirmed run successfully via SLURM batch jobs on Barkla —
prep (job 9762235) and baseline table (job 9762366) — on 12 July 2026.

---

# Steps 1-7 — Cohort construction (`prep_cohort.R`)

```{r prep-cohort}
#=====================================================================
# OPCRD External Validation — SHARED COHORT PREP
#=====================================================================

opcrd_path <- "/users/hlskhatt/data/OPCRD.csv"
pwp_path   <- "/users/hlskhatt/data/PWP asthma - overall,definite.xls"
output_dir <- "/users/hlskhatt/outputs"
if (!dir.exists(output_dir)) dir.create(output_dir, recursive = TRUE)

library(dplyr)
library(data.table)
library(moments)
library(lubridate)
library(ggplot2)

#=========================
# STEP 1: LOADING THE DATA
#=========================
opcrd <- as.data.frame(fread(opcrd_path))
cat("Rows loaded:", nrow(opcrd), "\n")
cat("Columns loaded:", ncol(opcrd), "\n")

# ========================================================
# STEP 2: COHORT DERIVATION — ACTIVE ASTHMA THERAPY FILTER
# ========================================================
defasthma <- subset(opcrd, !(ba_Drug_Treatment %in%
                                c("SAAAC + SABA", "SAAC", "SABA")))
cat("Unique patients after active-therapy filter:",
    length(unique(defasthma$Patient_ID)), "\n")

opcrda <- data.table(defasthma, key = "Patient_ID")
year1  <- opcrda[J(unique(Patient_ID)), mult = "first"]
cat("Final cohort size (year1):", nrow(year1), "\n")

#=======================================================
# STEP 3: RECODE MISSING VALUES FOR CONTINUOUS VARIABLES
#=======================================================
year1 <- year1 %>%
  mutate(
    BMI_clean = ifelse(BMI == -1, NA, BMI),
    Eosinophil_clean = ifelse(ba_Eosinophil_Count == -1,
                               NA, ba_Eosinophil_Count)
  )
cat("Missing BMI (after recode):", sum(is.na(year1$BMI_clean)), "\n")
cat("Missing Eosinophil (after recode):", sum(is.na(year1$Eosinophil_clean)), "\n")

#=========================================================
# STEP 4: DISTRIBUTION ASSESSMENT FOR CONTINUOUS VARIABLES
#=========================================================
continuous_vars <- c("age", "BMI_clean")
skewness_table <- data.frame(
  variable = continuous_vars,
  mean     = sapply(year1[, continuous_vars, with = FALSE], mean, na.rm = TRUE),
  median   = sapply(year1[, continuous_vars, with = FALSE], median, na.rm = TRUE),
  skewness = sapply(year1[, continuous_vars, with = FALSE], skewness, na.rm = TRUE)
)
print(skewness_table)
write.csv(skewness_table, file.path(output_dir, "skewness_table.csv"), row.names = FALSE)

age_hist <- ggplot(as.data.frame(year1), aes(x = age)) +
  geom_histogram(bins = 30, fill = "#7570b3", colour = "black") +
  theme_bw() + ggtitle("Distribution of age")
ggsave(file.path(output_dir, "age_distribution.png"), age_hist, width = 6, height = 4)

bmi_hist <- ggplot(as.data.frame(year1), aes(x = BMI_clean)) +
  geom_histogram(bins = 30, fill = "#1b9e77", colour = "black") +
  theme_bw() + ggtitle("Distribution of BMI")
ggsave(file.path(output_dir, "bmi_distribution.png"), bmi_hist, width = 6, height = 4)

#====================================
# STEP 5: LABEL CATEGORICAL VARIABLES
#====================================
year1 <- year1 %>%
  mutate(
    Gender_clean = factor(Gender_Coded,
                           levels = c(0, 1),
                           labels = c("Female", "Male")),
    Smoking_clean = ifelse(Smoking_Status == -1, NA, Smoking_Status),
    Smoking_clean = factor(Smoking_clean,
                            levels = c(1, 2, 3),
                            labels = c("Non-smoker",
                                       "Current smoker",
                                       "Ex-smoker"))
  )
print(table(year1$Gender_clean, useNA = "always"))
print(table(year1$Smoking_clean, useNA = "always"))

#=================================================================
# STEP 6: DERIVE PREVIOUS EXACERBATION VARIABLES AND YEAR OF ENTRY
#=================================================================
cat("Broad definition event count:",
    sum(year1$ba_Asthma_Exacerbations > 0), "\n")
cat("ATS-strict definition event count:",
    sum(year1$ba_Asthma_Exac_ATS > 0), "\n")

year1 <- year1 %>%
  mutate(
    prev_exac = factor(ifelse(ba_Asthma_Exacerbations > 0, "Yes", "No"),
                        levels = c("No", "Yes")),
    year_of_entry = factor(year(dmy(Start_Year_1)))
  )
cat("Missing year_of_entry (should be 0):", sum(is.na(year1$year_of_entry)), "\n")
print(table(year1$prev_exac))
print(table(year1$year_of_entry))

#=====================================================================
# STEP 7: LOAD PWP OUTCOME FILE AND DERIVE EXACERBATION STRATIFICATION
#=====================================================================
pwp <- read.delim(pwp_path, header = FALSE, sep = "\t",
                   col.names = c("Patient_ID", "tstart", "tstop", "event", "eventnum"))
cat("PWP rows:", nrow(pwp), "| columns:", ncol(pwp),
    "| unique patients:", length(unique(pwp$Patient_ID)), "\n")

exac_counts <- pwp %>%
  group_by(Patient_ID) %>%
  summarise(n_exac = sum(event == 1))

year1 <- year1 %>%
  left_join(exac_counts, by = "Patient_ID") %>%
  mutate(
    exac_group = case_when(
      is.na(n_exac) ~ "Outcome unknown",
      n_exac == 0   ~ "0",
      n_exac == 1   ~ "1",
      n_exac >= 2   ~ ">=2"
    ),
    exac_group = factor(exac_group,
                         levels = c("0", "1", ">=2", "Outcome unknown"))
  )
print(table(year1$exac_group))
cat("Total (should equal cohort size):", sum(table(year1$exac_group)), "\n")

saveRDS(year1, file.path(output_dir, "year1.rds"))
saveRDS(pwp,   file.path(output_dir, "pwp.rds"))

cat("\n--- PREP COMPLETE ---\n")
cat("Cohort size:", nrow(year1), "| PWP patients:", length(unique(pwp$Patient_ID)), "\n")
```

---

# Steps 8-9 — Table build and export (`baseline_table.R`)

```{r baseline-table}
#=====================================================================
# OPCRD External Validation — BASELINE CHARACTERISTICS TABLE (Table 1)
#=====================================================================

output_dir <- "/users/hlskhatt/outputs"

library(dplyr)
library(gtsummary)

year1 <- readRDS(file.path(output_dir, "year1.rds"))
cat("Loaded cohort size:", nrow(year1), "\n")

#======================================================
# STEP 8: BUILDING FINAL BASELINE CHARACTERISTICS TABLE
#======================================================
vars_to_summarise <- c("Gender_clean", "prev_exac", "Smoking_clean",
                        "year_of_entry", "age", "BMI_clean")

var_labels <- list(
  Gender_clean  ~ "Gender",
  prev_exac     ~ "Previous exacerbation (broad definition)",
  Smoking_clean ~ "Smoking status",
  year_of_entry ~ "Year of entry into study",
  age           ~ "Age (years)",
  BMI_clean     ~ "BMI"
)

overall_tbl <- as.data.frame(year1) %>%
  select(all_of(vars_to_summarise)) %>%
  tbl_summary(
    statistic = list(
      age       ~ "{mean} ({sd})",
      BMI_clean ~ "{median} ({p25}, {p75})",
      all_categorical() ~ "{n} ({p}%)"
    ),
    label        = var_labels,
    missing_text = "Unknown",
    type         = list(prev_exac ~ "categorical")
  )

strat_tbl <- as.data.frame(year1) %>%
  select(all_of(vars_to_summarise), exac_group) %>%
  tbl_summary(
    by      = exac_group,
    percent = "row",
    statistic = list(
      age       ~ "{mean} ({sd})",
      BMI_clean ~ "{median} ({p25}, {p75})",
      all_categorical() ~ "{n} ({p}%)"
    ),
    label        = var_labels,
    missing_text = "Unknown",
    type         = list(prev_exac ~ "categorical")
  )

table1_final <- tbl_merge(
  tbls        = list(overall_tbl, strat_tbl),
  tab_spanner = c("**Overall**",
                  "**Number of exacerbations, full study period (baseline + outcome years, days 0-1095)**")
) %>%
  bold_labels()

print(table1_final)

#======================================
# STEP 9: EXPORT TABLE TO WORD DOCUMENT
#======================================
table1_final %>%
  as_gt() %>%
  gt::gtsave(filename = file.path(output_dir, "Table1.docx"))

cat("\n--- BASELINE TABLE COMPLETE ---\n")
cat("Saved:", file.path(output_dir, "Table1.docx"), "\n")
```
