---
title: "External Validation of Noble et al. (2021)"
author: "Saad Ishaq Khattak"
date: "Confirmed running on Barkla HPC, 12 July 2026 (jobs 9763679, 9763685)"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(eval = FALSE, echo = TRUE)
```

# Purpose

External validation of the Noble et al. (2021) asthma exacerbation
prediction model in OPCRD. Loads `year1.rds` and `pwp.rds`, produced by
the shared prep script (`prep_cohort.R`) — this script does not rebuild
the cohort itself.

**Why Noble was chosen:** of the 18 models identified by the systematic
review, Noble was the strongest candidate for external validation — a
fully published equation with intercept, a UK primary care population
matching OPCRD's eligibility criteria, and 13 of 14 predictors directly
available.

**Why this validation is a worked example, not the dissertation's
primary contribution:** attempting it surfaced structural limitations
that no amount of careful coding can resolve — Noble's true outcome
(hospitalisation/A&E/death) does not exist in OPCRD and was substituted
using the PWP recurrent-event file; anaphylaxis history does not exist
in OPCRD at all and was set to zero for every patient; GINA management
step required inference via a differently-named treatment-intensity
variable. None of these are coding errors — they are properties of the
dataset, and they demonstrate empirically why classical external
validation of this literature is insufficient, motivating the
dissertation's actual centerpiece (IECV with Debray pooling).

**Key result:** the worked example check (reproducing Noble's own
published example patient, expected LP = -4.591) matched exactly,
confirming the 22-coefficient equation was transcribed correctly.
C-statistic 0.728 (0.723-0.733) closely matches Noble's own reported
0.71, but calibration is poor (O/E ratio 7.933) — the model
substantially under-predicts risk on this substitute outcome, the
expected signature of applying a model calibrated for a rare, severe
outcome onto a broader, more common one.

**Status:** confirmed run successfully via SLURM batch job on Barkla
(job 9763685, following an identical direct-run confirmation) on
12 July 2026. No patient data or output files are included in this
repository — code only. Outputs (comparison table, decile summary,
summary statistics) are saved on Barkla and the author's own machine,
not in this repository, per the data governance note in the main
README.

```{r noble-validation}
#=====================================================================
# OPCRD External Validation — NOBLE (2021) EXTERNAL VALIDATION
#=====================================================================

output_dir <- "/users/hlskhatt/outputs"

library(dplyr)
library(pROC)
library(pmvalsampsize)

year1 <- readRDS(file.path(output_dir, "year1.rds"))
pwp   <- readRDS(file.path(output_dir, "pwp.rds"))
cat("Loaded cohort size:", nrow(year1), "\n")
cat("Loaded PWP rows:", nrow(pwp), "\n")

#=============================================
# MISSINGNESS CHECK — DIRECT BINARY PREDICTORS
#=============================================
vars_to_check <- c("Rhinitis_Dx_drugs", "Diabetes_Dx", "IHD_Dx",
                    "Anxiety_depression_Dx", "Paracetamol",
                    "ba_Acute_OS_Courses", "ba_Antibiotics", "ba_Asthma_InP")

for (v in vars_to_check) {
  cat("\n---", v, "---\n")
  print(table(year1[[v]], useNA = "always"))
}

neg1_flags <- sapply(vars_to_check, function(v) any(year1[[v]] == -1, na.rm = TRUE))
neg1_counts <- sapply(vars_to_check, function(v) sum(year1[[v]] == -1, na.rm = TRUE))
cat("\nVariables containing -1:\n")
print(neg1_flags)
cat("\nCount of -1 values per variable:\n")
print(neg1_counts)

#=========================================================
# STEP 10: MAP AND CONSTRUCT NOBLE (2021) MODEL PREDICTORS
#=========================================================
year1 <- year1 %>%
  mutate(
    age_61_80 = ifelse(age >= 61, 1, 0),
    BMI_underweight = ifelse(BMI_clean < 18.5, 1, 0),
    BMI_overweight  = ifelse(BMI_clean >= 25 & BMI_clean < 30, 1, 0),
    BMI_obese       = ifelse(BMI_clean >= 30, 1, 0),
    smoker_current = ifelse(Smoking_clean == "Current smoker", 1, 0),
    smoker_ex      = ifelse(Smoking_clean == "Ex-smoker", 1, 0),
    eos_over400 = ifelse(Eosinophil_clean > 0.4, 1, 0),
    rhinitis     = Rhinitis_Dx_drugs,
    diabetes     = Diabetes_Dx,
    ihd          = IHD_Dx,
    anx_dep      = Anxiety_depression_Dx,
    paracetamol  = Paracetamol,
    anaphylaxis = 0,
    step_therapy_clean = ifelse(ba_step_therapy == -1, NA, ba_step_therapy),
    gina_notherapy = ifelse(step_therapy_clean == 0, 1, 0),
    gina_step3     = ifelse(step_therapy_clean == 3, 1, 0),
    gina_step4_5   = ifelse(step_therapy_clean %in% c(4, 5), 1, 0),
    lrti_1  = ifelse(ba_Antibiotics == 1, 1, 0),
    lrti_2p = ifelse(ba_Antibiotics >= 2, 1, 0),
    ocs_1  = ifelse(ba_Acute_OS_Courses == 1, 1, 0),
    ocs_2  = ifelse(ba_Acute_OS_Courses == 2, 1, 0),
    ocs_3p = ifelse(ba_Acute_OS_Courses >= 3, 1, 0),
    prior_hosp = ifelse(ba_Asthma_InP >= 1, 1, 0)
  )
cat("Row count after Step 10:", nrow(year1), "\n")

#=============================================================
# STEP 11: CONSTRUCTING NOBLE OUTCOME SUBSTITUTE FROM PWP FILE
#=============================================================
pwp_365_v2 <- pwp %>%
  filter(tstart >= 365 & tstop <= 730 & event == 1)

noble_outcome_flag_v2 <- pwp_365_v2 %>%
  group_by(Patient_ID) %>%
  summarise(noble_event_365_v2 = 1) %>%
  ungroup()

year1 <- year1 %>%
  left_join(noble_outcome_flag_v2, by = "Patient_ID") %>%
  mutate(
    noble_outcome_v2 = case_when(
      !is.na(noble_event_365_v2)             ~ 1,
      Patient_ID %in% unique(pwp$Patient_ID)  ~ 0,
      TRUE                                    ~ NA_real_
    )
  )

print(table(year1$noble_outcome_v2, useNA = "always"))
cat("Row count check:", nrow(year1) == 155163, "\n")
cat("N with known outcome:", sum(!is.na(year1$noble_outcome_v2)), "\n")

#===========================================================
# STEP 12: CALCULATING OUTCOME PREVALENCE ON THE FULL COHORT
#===========================================================
noble_prevalence <- mean(year1$noble_outcome_v2, na.rm = TRUE)
cat("\nNoble-substitute outcome prevalence:", noble_prevalence, "\n")

#=================================================================
# STEP 13: CALCULATING THE LINEAR PREDICTOR (LP) FOR EVERY PATIENT
#=================================================================
year1 <- year1 %>%
  mutate(
    noble_LP =
      -5.013 +
      0.192 * age_61_80 +
      0.597 * BMI_underweight +
      -0.210 * BMI_overweight +
      -0.103 * BMI_obese +
      0.251 * smoker_current +
      0.093 * smoker_ex +
      0.237 * eos_over400 +
      -0.212 * rhinitis +
      0.378 * diabetes +
      0.243 * ihd +
      0.192 * anx_dep +
      0.790 * anaphylaxis +
      0.568 * gina_notherapy +
      0.101 * gina_step3 +
      0.461 * gina_step4_5 +
      0.313 * lrti_1 +
      0.206 * lrti_2p +
      0.551 * ocs_1 +
      0.975 * ocs_2 +
      1.141 * ocs_3p +
      0.204 * paracetamol +
      1.877 * prior_hosp,
    noble_pred_prob = exp(noble_LP) / (1 + exp(noble_LP))
  )

cat("Row count after Step 13:", nrow(year1), "\n")
cat("N with non-missing LP:", sum(!is.na(year1$noble_LP)), "\n")

worked_example <- data.frame(
  age_61_80 = 0, BMI_underweight = 0, BMI_overweight = 1, BMI_obese = 0,
  smoker_current = 0, smoker_ex = 0, eos_over400 = 0,
  rhinitis = 1, diabetes = 0, ihd = 0, anx_dep = 1, anaphylaxis = 0,
  gina_notherapy = 0, gina_step3 = 1, gina_step4_5 = 0,
  lrti_1 = 0, lrti_2p = 0, ocs_1 = 1, ocs_2 = 0, ocs_3p = 0,
  paracetamol = 0, prior_hosp = 0
)
worked_LP <- with(worked_example,
  -5.013 + 0.192*age_61_80 + 0.597*BMI_underweight - 0.210*BMI_overweight -
  0.103*BMI_obese + 0.251*smoker_current + 0.093*smoker_ex +
  0.237*eos_over400 - 0.212*rhinitis + 0.378*diabetes + 0.243*ihd +
  0.192*anx_dep + 0.790*anaphylaxis + 0.568*gina_notherapy +
  0.101*gina_step3 + 0.461*gina_step4_5 + 0.313*lrti_1 + 0.206*lrti_2p +
  0.551*ocs_1 + 0.975*ocs_2 + 1.141*ocs_3p + 0.204*paracetamol +
  1.877*prior_hosp
)
worked_prob <- exp(worked_LP) / (1 + exp(worked_LP))
cat("\nWorked example LP -- expected -4.591, got:", round(worked_LP, 3), "\n")
cat("MATCH:", round(worked_LP, 3) == -4.591, "\n")
worked_example_match <- round(worked_LP, 3) == -4.591

#====================================
# STEP 14: SD OF THE LINEAR PREDICTOR
#====================================
noble_LP_sd <- sd(year1$noble_LP, na.rm = TRUE)
noble_LP_mean <- mean(year1$noble_LP, na.rm = TRUE)
cat("\nSD of Noble linear predictor:", noble_LP_sd, "\n")

#===================================================================
# STEP 15: pmvalsampsize -- SAMPLE SIZE ADEQUACY CHECK
#===================================================================
noble_samplesize <- pmvalsampsize(
  type = "b", cstatistic = 0.71, oe = 1, cslope = 1,
  prevalence = 0.0964248, lpnormal = c(-4.436176, 0.5769082)
)
print(noble_samplesize)

#==============================================================
# STEP 16: SUBSET TO COMPLETE CASES — THE TRUE ANALYSIS DATASET
#==============================================================
noble_complete_v2 <- year1 %>%
  filter(!is.na(noble_LP) & !is.na(noble_outcome_v2))

cat("TRUE complete-case N:", nrow(noble_complete_v2), "\n")

true_prevalence <- mean(noble_complete_v2$noble_outcome_v2)
true_LP_mean    <- mean(noble_complete_v2$noble_LP)
true_LP_sd      <- sd(noble_complete_v2$noble_LP)

noble_samplesize_true <- pmvalsampsize(
  type = "b", cstatistic = 0.71, oe = 1, cslope = 1,
  prevalence = true_prevalence, lpnormal = c(true_LP_mean, true_LP_sd)
)
print(noble_samplesize_true)

true_adequacy <- nrow(noble_complete_v2) / noble_samplesize_true$sample_size
cat("\nTRUE adequacy multiple:", round(true_adequacy, 2), "\n")

#==============================================================
# STEP 18 / 18B: PERFORMANCE MEASURES (LOGISTIC + DECILE METHODS)
#==============================================================
noble_roc_v2 <- roc(noble_complete_v2$noble_outcome_v2, noble_complete_v2$noble_pred_prob,
                     ci = TRUE, ci.method = "delong")
cat("\nC-statistic:", round(as.numeric(noble_roc_v2$auc), 3), "\n")

citl_model_v2 <- glm(noble_outcome_v2 ~ offset(noble_LP),
                      data = noble_complete_v2, family = binomial)
citl_estimate_v2 <- coef(citl_model_v2)[1]
citl_ci_v2 <- confint(citl_model_v2)

observed_v2 <- sum(noble_complete_v2$noble_outcome_v2)
expected_v2 <- sum(noble_complete_v2$noble_pred_prob)
oe_ratio_v2 <- observed_v2 / expected_v2
oe_se_log_v2 <- sqrt(1 / observed_v2)
oe_ci_lower_v2 <- exp(log(oe_ratio_v2) - 1.96 * oe_se_log_v2)
oe_ci_upper_v2 <- exp(log(oe_ratio_v2) + 1.96 * oe_se_log_v2)

slope_model_v2 <- glm(noble_outcome_v2 ~ noble_LP,
                       data = noble_complete_v2, family = binomial)
slope_estimate_v2 <- coef(slope_model_v2)["noble_LP"]
slope_ci_v2 <- confint(slope_model_v2)["noble_LP", ]

noble_complete_v2 <- noble_complete_v2 %>%
  mutate(risk_decile = ntile(noble_pred_prob, 10))

decile_summary_v2 <- noble_complete_v2 %>%
  group_by(risk_decile) %>%
  summarise(
    n_patients      = n(),
    mean_pred_risk  = mean(noble_pred_prob),
    observed_risk   = mean(noble_outcome_v2)
  ) %>%
  ungroup()
print(decile_summary_v2)

decile_slope_model_v2 <- lm(observed_risk ~ mean_pred_risk, data = decile_summary_v2)
decile_slope_estimate_v2 <- coef(decile_slope_model_v2)["mean_pred_risk"]
decile_slope_ci_v2 <- confint(decile_slope_model_v2)["mean_pred_risk", ]

#===================================================================
# STEP 19: FORMAL COMPARISON TO NOBLE'S REPORTED VALIDATION FIGURES
#===================================================================
comparison_table_v2 <- data.frame(
  Measure = c("C-statistic",
              "Calibration-in-the-large (CITL)",
              "O/E ratio",
              "Calibration slope (logistic method)",
              "Calibration slope (decile method)"),
  Noble_reported = c("0.71 [0.70-0.72]", "Not reported", "Not reported",
                      "0.85 [0.75-0.96]", "0.85 [0.75-0.96]"),
  This_validation = c(
    paste0(round(as.numeric(noble_roc_v2$auc),3), " [",
           round(noble_roc_v2$ci[1],3), "-", round(noble_roc_v2$ci[3],3), "]"),
    paste0(round(citl_estimate_v2,3), " [",
           round(citl_ci_v2[1],3), "-", round(citl_ci_v2[2],3), "]"),
    paste0(round(oe_ratio_v2,3), " [",
           round(oe_ci_lower_v2,3), "-", round(oe_ci_upper_v2,3), "]"),
    paste0(round(slope_estimate_v2,3), " [",
           round(slope_ci_v2[1],3), "-", round(slope_ci_v2[2],3), "]"),
    paste0(round(decile_slope_estimate_v2,3), " [",
           round(decile_slope_ci_v2[1],3), "-", round(decile_slope_ci_v2[2],3), "]")
  )
)
print(comparison_table_v2, row.names = FALSE)

#=====================================================================
# SAVE OUTPUTS — everything needed for the write-up, in one place
#=====================================================================
write.csv(comparison_table_v2, file.path(output_dir, "Noble_comparison_table.csv"), row.names = FALSE)
write.csv(decile_summary_v2, file.path(output_dir, "Noble_decile_summary.csv"), row.names = FALSE)

summary_stats <- data.frame(
  Item = c("Worked example LP matches paper (-4.591)",
           "Full cohort N", "PWP patients (known outcome)",
           "Complete-case analysis N (predictors + outcome)",
           "Outcome prevalence, full cohort", "Outcome prevalence, complete-case",
           "Sample size required (pmvalsampsize)", "Adequacy multiple",
           "C-statistic", "CITL", "O/E ratio",
           "Calibration slope (logistic)", "Calibration slope (decile)"),
  Value = c(worked_example_match,
            nrow(year1), sum(!is.na(year1$noble_outcome_v2)),
            nrow(noble_complete_v2),
            round(noble_prevalence, 4), round(true_prevalence, 4),
            noble_samplesize_true$sample_size, round(true_adequacy, 2),
            round(as.numeric(noble_roc_v2$auc), 3), round(citl_estimate_v2, 3),
            round(oe_ratio_v2, 3), round(slope_estimate_v2, 3),
            round(decile_slope_estimate_v2, 3))
)
write.csv(summary_stats, file.path(output_dir, "Noble_summary_stats.csv"), row.names = FALSE)

cat("\n--- NOBLE VALIDATION COMPLETE ---\n")
```
